### 2014-03-30 paper_enahpc14/Introduction.tex
updates introduction and bibliography

The  interaction  between aerosols  and  clouds  presents  one of  the
biggest  uncertainties  in   present  day  climate  simulations.   The
chemistry involved is complex,  but there is a concerted international
effort to model the underlying processes through models. These models,
such as  the Aerosol  Reactive Transport (ART),  developed at  the KIT
(Karlsruhe  Institute of  Technology)  , offer  a  key opportunity  to
reduce   the  climate  uncertainty,   particularly  on   the  regional
scale. These  models can be coupled  with climate models,  such as the
regional  weather  and   climate  COSMO  (Consortium  for  Small-scale
Modelling) model  developed by the German Weather  Service (DWD).  The
COSMO model  is a weather forecast  model which has  become a standard
all  over Europe.   Beside the  federal weather  forecast  stations in
Germany  (Deutscher  Wetterdienst,  DWD),  Switzerland  (MeteoSchweiz,
MCH),  Italy (Ufficio  Generale  Spazio Aereo  e Meteorologia,  USAM),
Greece  (Hellenic   National  Meteorological  Service,   HNMS)  Poland
(Institute  of   Meteorology  and  Water   Management,  IMGW)  Romania
(National  Meteorological  Administration,  NMA) and  Russia  (Federal
Service for Hydrometeorology  and Environmental Monitoring, RHM), also
a   large  number   of  agencies   including  military   and  research
institutions  base their  forecasts on  COSMO. The  combined COSMO-ART
model is capable of simulating  aerosol distributions as well as their
interactions over Europe as well as other regional domains.

It has  been a major  scientific achievement to effectively  model the
underlying  chemical  processes   through  these  applications.   This
achievement  now   needs  to  be  fostered  by   reducing  its  energy
expense. Currently, resource allocation  is one of the key limitations
to scientists  in running the simulation periods  and resolutions they
would like.

>>>>

Comprehensive  scientific assessments  of climate-warming  trends over
the  past 50  years have  shown  that the  state and  dynamics of  the
Earth's climate system have  undergone unprecedented major changes, as
evidenced  by  increases  in   global  average  atmosphere  and  ocean
temperatures,  widespread melting  of  snow and  ice,  and rising  sea
levels.   The last report  of the  Intergovernmental Panel  on Climate
Change, released  in September 2013  \citep{IPCC-2013}, concluded that
most of the observed temperature  increases since the mid 20th Century
have  been caused  by  increasing concentrations  of greenhouse  gases
resulting from  anthropogenic activities  such as fossil  fuel burning
and  deforestation.  Long-lived greenhouse  gases, for  example carbon
dioxide, methane  and nitrous oxide, are chemically  stable and linger
homogeneously well-mixed and over time scales of a decade to thousands
of years in  the atmosphere; whereas short-lived gases  such as water,
suphur  dioxide  and  carbon  monoxide  which  respond  physically  or
chemically  to changes  in temperature,  act primarily  as  a feedback
mechanism.   Some   extremly  powerful  greenhouse   gases,  such  as
fluorinated gases can even have  a warming effect on the atmosphere up
to  23000  times  greater  than  carbon  dioxide  \citep{Zehner-2012}.
Fortunately  the  European  Parliament,  which  has  gained  worldwide
recognition as a  leader in climate policy, has  passed legislation in
March 2014  to phase  out fluorinated gas  emissions by  two-thirds by
2030,  towards  the  internationally  agreed goal  of  keeping  global
warming  below  2  degrees  Celsius  compared to  the  temperature  in
pre-industrial times.\\

\noindent
Current  cumulative releases of  human-induced gas  emissions enhances
the greenhouse  effect to  the Earth's atmosphere  and thus  creates a
large imbalance between incoming solar absorbed radiation and outgoing
longwave radiation emitted back  to space, through radiative trapping.
While the  Earth's temperature  is dependent upon  the greenhouse-like
action of  the atmosphere, the  Earth's radiation balance  is strongly
influenced by several  other factors such as the  type of surface that
sunlight first  encounters.  Forests, grasslands,  ocean surfaces, ice
caps, deserts,  and cities all absorb, reflect,  and radiate radiation
differently.   Sunlight falling  on a  white glacier  surface strongly
reflects back into space, resulting  in minimal heating of the surface
and  lower atmosphere.   Sunlight falling  on  a dark  desert soil  is
strongly absorbed,  on the other hand, and  contributes to significant
heating of the surface and lower atmosphere.  Cloud cover also affects
greenhouse  warming by  both reducing  the amount  of  solar radiation
reaching the earth's  surface and by reducing the  amount of radiation
energy emitted into space.\\

\noindent 
Aerosols are  suspensions of liquid,  solid or mixed particles  in the
air  (including sea  salt,  mineral dust,  sulphate, nitrate,  organic
carbon and  black carbon),  with highly variable  chemical composition
and  size  distribution \citep{Putaud-2010}.   Although  they are  not
considered a heat-trapping greenhouse gas and have shorter atmospheric
lifetimes, they significantly affect  the Earth's radiation budget and
are acknowledged as one of  the most significant and uncertain aspects
in  anthropogenic forcing  over the  last 150  years \citep{Koch-2009,
IPCC-2013}.  Enhanced  aerosols concentrations can  impact the climate
system by  reflecting (e.g.  pure sulfates and  nitrates) or absorbing
(e.g.  black  carbon) solar radiation  and thereby exert a  cooling or
warming  effect on  the Earth-atmosphere  system, causing  a so-called
direct    radiative   forcing    \citep{Charlson-1991,   Haywood-2000,
Ramanathan-2001,  Liao-2005, Bangert-2012,  Lundgren-2013}.  Depending
on their  size and  chemical composition, they  can also act  as cloud
condensation  and   ice  nuclei   and  profoundly  impact   the  cloud
microphysical   processes  and   optical   properties  (cloud   albedo
effect,   \citealp{Twomey-1977}),  leading   to  changes   in  droplet
concentrations                \citep{Albrecht-1989}                and
precipitation   \citep{Rosenfeld-2000,   Khain-2008,  Pruppacher-2010,
Seifert-2012, Tao-2012, Lee-2012},  and therefore producing a negative
indirect   radiative    forcing   \citep{Haywood-2000,   Lohmann-2005,
VandenHeever-2011,   Rosenfeld-2013}.    However,   the   effects   of
anthropogenic aerosols on clouds and the hydrological cycle as well as
the  cloud   lifetime  effect  are  especially  hard   to  assess  and
quantify   \citep{IPCC-2013},   and   remain   one  of   the   largest
uncertainties in climate modeling and in climate change prediction due
to the lack of  understanding of cloud feedbacks \citep{Sherwood-2013,
Rosenfeld-2013, Lee-2013}. \\

\noindent 
Hence to  enhance our understanding of  aerosol-cloud interactions and
reduce  uncertainties  of aerosol  effects  in  climate, the  research
community is making a  concerted international effort to represent the
underlying chemical  processes through models.  These  models, such as
ART (Aerosols  and Reactive Trace  gases) extension, developed  at the
Karlsruhe Institute  of Technology (KIT),  offer a key  opportunity to
reduce   the  climate  uncertainty,   particularly  on   the  regional
scale \citep{Knote-2011, Bangert-2011,  Knote-2013}.  These models can
be coupled with climate models,  such as the regional weather forecast
model COSMO (Consortium  for Small-scale Modelling), jointly developed
by  a consortium  of  European weather  centers  including the  German
weather service  DWD and MeteoSwiss,  and used in the  climate version
COSMO-CLM by a wide  research community.  The extended COSMO-ART model
provides a detailed description of air pollution chemistry and aerosol
processes, and  is mainly  designed to study  air quality  and aerosol
meteorology feedbacks on short, episodic to annual time scales.  It is
capable  of   simulating  aerosol  distributions  as   well  as  their
interactions over Europe as well as other regional domains.\\

\noindent 
COSMO-ART is  computationally much more demanding than  the COSMO core
since a  large number of additional  tracers and processes  have to be
considered.  Thus this model is currently severely-limited in terms of
applicability and expensive in terms of energy consumption.  COSMO has
recently  been  ported  to  GPUs  within the  framework  of  the  High
Performance  and  High  Productivity  Computing (HP2C)  Initiative  to
optimize it  for computational and energy  efficiency.  Although these
developments will  facilitate the  application of COSMO  for numerical
weather prediction and climate  simulations, they do little to address
the  coupling with  the  ART model  extension,  for which  significant
investments  are still required  to take  it to  a similar  level. The
efficiency of ART is being  addressed in the EU Exa2Green project.  In
Sec.~\ref{sec:1},  we   introduce  the  aerosol  model   and  its  its
integration  into  COSMO.    Sec.~\ref{sec:2}  presents  the  hardware
considerations.  In  Sec.~\ref{sec:3} we describe  measurement devices
and energy-aware techniques employed  to evaluate the energy footprint
of the  considered application.  Finally,  Sec.~\ref{sec:4} presents a
benchmark  analysis  of  COSMO-ART   from  the  Exa2Green  project  to
highlight the critical components and an outline of the next steps for
model optimization.  The ultimate goal  of the project is to deliver a
prototype code, which  provides an energy efficiency of  at least five
times of the  baseline value.  Such an implementation  would allow the
community to  investigate critical questions at  higher resolution and
over longer periods, at reduced cost to the environment.
